FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2025-01-29 manuscript.Rmd
update 2025-01-29 13:15:59.904362

The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to the widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
<!-- CJ: Maybe we should give an example here. Does anyone know one? -->
<!-- NvD: Harris 1976 has a nice example about how cognitive dissonance theory can be formalized in multiple ways, which offer hypotheses that contradict each other -->

@@

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and theory improvement have been largely overlooked. <!-- NvD: In philosophy of science, induction is considered to be problematic (logically invalid). It is more appropriate to talk about abduction when it comes to theory development. I am fine with sticking to induction if that is what psychologists "know". In that case, I suggest adding a clarifying footnote on abduction vs induction.  -->
The present paper restores balance by applying, for the first time, <!-- NvD: Are we sure that this is the first time? Do we need to make this claim? -->
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing the concept of *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

In a progressive research program [@lakatosHistoryScienceIts1971],
<!-- NvD: This is not exactly what Lakatos said. Also, I don't think many psychologists are familiar with his work. We do not need to reference his perspective. -->
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
<!-- by successively increasing the empirical content of a theory. -->
<!-- In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
<!-- CJ: I am not sure if I agree; I think increasing the empirical content rather implies more and more precision, which loses generality but gains specificity. Not sure how to resolve.
This needs more work! -->
<!-- NvD: SOmething like? "In the envisioned ideal, the empirical cycle is a virtuous spiral where theories become increasingly more precise and/or universal (i.e., Popper's empirical content), offering hypotheses that can be more easily falsified / severely tested, data become more informative / less ambiguous about the patterns in nature, which in turn inform even more precise and/or universal theories. -->
.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature: 
According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
In addition, the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
The remaining 85% of deductive studies lacked an explicit connection between theory and hypothesis.
<!-- NvD: (most of) The other papers also don't have a true derivation chain. It is rare to see a logical derivation, analytical proof, or simulation that connects hypothesis to theory.  -->
In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020], and researchers are simply testing them as part of a cultural ritual [@gigerenzerNullRitualWhat2004].
<!-- NvD: I am not sure what is being said here. In the statistical sense, people are testing the null hypothesis. However, that is (almost always) not the hypothesis they are interested in nor the one they mention the introduction of their papers. Also, I thought that gigerenzer's paper was about the ritualistic use of NHST, not the origin of the hypotheses that were being tested. -->
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
<!-- NvD: I don't think that statement completely covers it. How about: "Testing ad-hoc hypotheses not grounded in theory might inform us of the psychological phenomena that we can consistently observe (e.g., cognitive biases, attentional blindess, comorbidity of depression and anxiety disorder, etc.). However, it does not advance our understanding of the processes and mechanisms that bring about these phenomena. -->
Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021].

@@

Aside from technical solutions, version control is a social process as well.
On the one hand, regular updates can improve theories - but on the other hand, it risks breaking the compatibility between theories and hypotheses derived from them, or compatibility between one theory and others that depend upon it.
For example, if we construct a theory to explain a specific phenomenon, and we cross-reference an existing theory comprising an ontology for our field - that dependency is broken if the ontology is later updated and our phenomenon of interest is removed.
<!-- NvD: Maybe we can add an  example (e.g., the theory that explains ego-depletion and the failed replication of the phenomenon). -->
In computer science, these challenges are navigated by assigning version numbers.
Specifically, *semantic versioning* comprises a simple set of rules for assigning version numbers to information artifacts.
Whereas version control tracks changes,
semantic versioning communicates what those changes mean to users of the theory.
We propose the following adaptation of semantic versioning for theories:

>>>>

The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
<!-- CJ: Maybe we should give an example here. Does anyone know one? -->
<!-- NvD: Harris 1976 has a nice example about how cognitive dissonance theory can be formalized in multiple ways, which offer hypotheses that contradict each other -->

@@

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and theory improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories; others have coined *abduction* to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960]. However, Peirce also acknowledged that abduction "is not essentially different from induction" (p.4881). Defining what makes a theory "best" requires making auxiliary assumptions. For present purposes, induction and abduction are interchangeable terms.].
<!-- Peirce: CP 8.233 Cross-Ref:†† -->
<!-- 233. This is not essentially different from induction. Only it is most -->
<!-- usually an induction from instances which are not discrete and numerable. I now -->
<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
<!-- line of procedure, or at least not to distinguish it sharply. -->
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing the concept of *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

In a progressive research program [@lakatosHistoryScienceIts1971],
<!-- NvD: This is not exactly what Lakatos said. Also, I don't think many psychologists are familiar with his work. We do not need to reference his perspective. -->
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
<!-- by successively increasing the empirical content of a theory. -->
<!-- In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
<!-- CJ: I am not sure if I agree; I think increasing the empirical content rather implies more and more precision, which loses generality but gains specificity. Not sure how to resolve.
This needs more work! -->
<!-- NvD: SOmething like? "In the envisioned ideal, the empirical cycle is a virtuous spiral where theories become increasingly more precise and/or universal (i.e., Popper's empirical content), offering hypotheses that can be more easily falsified / severely tested, data become more informative / less ambiguous about the patterns in nature, which in turn inform even more precise and/or universal theories. -->
.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature: 
According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
The remaining 85% of deductive studies lacked an explicit connection between theory and hypothesis.
<!-- NvD: (most of) The other papers also don't have a true derivation chain. It is rare to see a logical derivation, analytical proof, or simulation that connects hypothesis to theory.  -->
In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020], and researchers are simply testing them as part of a cultural ritual [@gigerenzerNullRitualWhat2004].
<!-- NvD: I am not sure what is being said here. In the statistical sense, people are testing the null hypothesis. However, that is (almost always) not the hypothesis they are interested in nor the one they mention the introduction of their papers. Also, I thought that gigerenzer's paper was about the ritualistic use of NHST, not the origin of the hypotheses that were being tested. -->
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
<!-- NvD: I don't think that statement completely covers it. How about: "Testing ad-hoc hypotheses not grounded in theory might inform us of the psychological phenomena that we can consistently observe (e.g., cognitive biases, attentional blindess, comorbidity of depression and anxiety disorder, etc.). However, it does not advance our understanding of the processes and mechanisms that bring about these phenomena. -->
Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021].

@@

Aside from technical solutions, version control is a social process as well.
On the one hand, regular updates can improve theories - but on the other hand, it risks breaking compatibility between theories and hypotheses derived from them, or compatibility between one theory and others that depend upon it.
For example, if we construct a theory to explain a specific phenomenon, and we cross-reference an existing theory comprising an ontology for our field - that dependency is broken if the ontology is later updated and our phenomenon of interest is removed.
<!-- NvD: Maybe we can add an  example (e.g., the theory that explains ego-depletion and the failed replication of the phenomenon). -->
In computer science, these challenges are navigated by assigning version numbers.
Specifically, *semantic versioning* comprises a simple set of rules for assigning version numbers to information artifacts.
Whereas version control tracks changes,
semantic versioning communicates what those changes mean to users of the theory.
We propose the following adaptation of semantic versioning for theories:
